---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r}
# Create the data folder if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Define the file path to the plots directory
plots_dir <- "plots"

# Create the plots folder if it doesn't exist
if (!dir.exists(plots_dir)) {
  dir.create(plots_dir)
}

# Define the file path to the results directory
results_dir <- "results"

# Create the results folder if it doesn't exist
if (!dir.exists(results_dir)) {
  dir.create(results_dir)
}
```


```{r}


#contains DEGs subset from DESeq2
data_dir <- file.path("data")
data_file <- file.path(data_dir, "DGES_Fold_Change_Rice_800.csv")
metadata_file <- file.path(data_dir, "Col_info_Rice.csv")




```

Now that our file paths are declared, we can use the `file.exists()` function to check that the files are where we specified above. 

```{r}
# Check if the gene expression matrix file is at the path stored in `data_file`
file.exists(data_file)

# Check if the metadata file is at the file path stored in `metadata_file`
file.exists(metadata_file)
```
***

<!-- Do not delete this line --> <a name="analysis" style="padding-top:56px;margin-top:-56px;">&nbsp;</a>

# Identifying co-expression gene modules with WGCNA - RNA-seq

## Install libraries

```{r}
if (!("pacman" %in% installed.packages())) {
  install.packages("pacman", dependencies = TRUE)
}
pacman::p_load(BiocManager, magrittr, DESeq2, ggplot2, impute, WGCNA, ggforce, ComplexHeatmap, install = TRUE)

#re using R-3.6.0, but trying to install Bioconductor version 3.8. Try using BiocManager::install("Biobase"). If that works, then try BiocManager::install("GO.db").

#If you have success, make sure your installation is valid with BiocManager::valid().

```


## Import and set up data

```{r}
# Read in metadata TSV file
metadata <- readr::read_csv(metadata_file)
metadata.sub <- metadata %>% subset(., select = c(2,4,5)) %>% dplyr::distinct(., sample, .keep_all = TRUE) 

# Read in data TSV file
df <- readr::read_csv(data_file) %>%
  # Here we are going to store the gene IDs as row names so that we can have a numeric matrix to perform calculations on later
  tibble::column_to_rownames("gene")

#sorghum.samples <- c("A1_S73_L001","A1_S73_L002","A2_S74_L001","A2_S74_L002","A3_S75_L001","A3_S75_L002","A5_S76_L001","A5_S76_L002","A6_S77_L001","A6_S77_L002","A7_S777_L001","A7_S777_L002","AC1_S78_L001","AC1_S78_L002","AC2_S79_L001","AC2_S79_L002","AC3_S80_L001","AC3_S80_L002","B1_S81_L001","B1_S81_L002","B2_S82_L001","B2_S82_L002","B3_S83_L001","B3_S83_L002","B5_S84_L001","B5_S84_L002","B6_S85_L001","B6_S85_L002","B7_S86_L001","B7_S86_L002","BC1_S87_L001","BC1_S87_L002","BC2_S88_L001","BC2_S88_L002","BC3_S89_L001","BC3_S89_L002","C1_S90_L001","C1_S90_L002","C2_S91_L001","C2_S91_L002","C3_S92_L001","C3_S92_L002","C6_S93_L001","C6_S93_L002","C7_S94_L001","C7_S94_L002","C8_S95_L001","C8_S95_L002","CC1_S96_L001","CC1_S96_L002","CC2_S97_L001","CC2_S97_L002","CC3_S98_L001","CC3_S98_L002","D1_S99_L001","D1_S99_L002","D2_S100_L001","D2_S100_L002","D3_S101_L001","D3_S101_L002","D5_S102_L001","D5_S102_L002","D7_S103_L001","D7_S103_L002","D8_S104_L001","D8_S104_L002","DC1_S105_L001","DC1_S105_L002","DC2_S106_L001","DC2_S106_L002","DC3_S107_L001","DC3_S107_L002","E1_S108_L001","E1_S108_L002","E2_S109_L001","E2_S109_L002","E4_S110_L001","E4_S110_L002","E5_S111_L001","E5_S111_L002","E6_S112_L001","E6_S112_L002","E7_S113_L001","E7_S113_L002","EC1_S114_L001","EC1_S114_L002","EC2_S115_L001","EC2_S115_L002","EC3_S116_L001","EC3_S116_L002")

counts.rice <- df
names(metadata) <- c("sample_id", "variety","times","treatment", "sample")
metadata <- metadata %>% dplyr::filter(., sample_id %in% colnames(counts.rice))

```

## Check sample names bewteen counts and experimental setup
Reorder the data and perform the checks.
If the output to any of the 2 checks below is FALSE, then recheck the data

```{r}
names <- as.character(metadata$sample_id)
counts.rice <- subset(counts.rice, select = names) ## reorder counts columns based on the experimental setup 
dim(counts.rice)
dim(metadata)

## check the results
# is.null(setdiff(colnames(counts), metadata$Sample_Name))
all(colnames(counts.rice) %in% metadata$sample_id)
all(colnames(counts.rice) == metadata$sample_id)

m <- data.matrix(counts.rice[, 1:ncol(counts.rice)])# convert readcount table to matrix
tibble::tibble(metadata) %>% 
  rmarkdown::paged_table(.)

```


## Raw counts are converted to CPM to log-CPM

```{r}
## Calculate CPM from raw counts to normalize for different sequencing depth for each sample
cpm <- apply(m,2, function(x) (x/sum(x))*1000000) 

## remove any genes with a CPM of less than 1 across all samples
m1 <- m[rowSums(cpm < 1) < 37, ]  
```

# Counts to keep
```{r}
options(scipen = 100)
thresh <- m1 > 1 ###to confirm the output of filtering
# head(thresh)
table(rowSums(thresh))## there are 7847 genes that have TRUES in all 174 samples
keep <- rowSums(thresh) >= 30 ## keep genes that have at least 30 TRUES in each row of thresh
counts.keep <- m1[keep,] ## subset the rows of count-data to keep the more highly expressed genes
summary(keep)
dim(counts.keep)
plot(m1[,5],m1[,5])

```

# DESeq DGE analysis

Using the <a href="http://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html" target="_blank">DESeq protocol</a> We analysed the data through using DESeq to see if we can replicate the DEG's identified using limma-voom

## Create DESeq object

```{r}
metadata$sample <- factor(metadata$sample)
metadata$times <- factor(metadata$times)
metadata$sample_id <- factor(metadata$sample_id)
metadata$variety <- factor(metadata$variety)

dds.raw <- DESeqDataSetFromMatrix(countData = counts.keep, colData = metadata, design = ~0+variety)

## Collapse Batches for each sample
dds <- collapseReplicates(dds.raw, dds.raw$sample, dds.raw$sample_id)
```

## Format normalized data for WGCNA

Extract the normalized counts to a matrix and transpose it so we can pass it to WGCNA.

```{r}
# Normalize and transform the data in the `DESeqDataSet` object using the `vst()`
# function from the `DESEq2` R package
#transform dds
#dds_norm <- vst(dds)
#normalized_counts <- assay(dds_norm) %>%
  #t() # Transpose this data

dds_norm <- vst(dds, blind = TRUE, nsub = nrow(dds))
normalized_counts <- assay(dds_norm) %>%
  t() # Transpose this data

```
## Determine parameters for WGCNA

To identify which genes are in the same modules, WGCNA first creates a weighted network to define which genes are near each other. 
The measure of "adjacency" it uses is based on the correlation matrix, but requires the definition of a threshold value, which in turn depends on a "power" parameter that defines the exponent used when transforming the correlation values. 
The choice of power parameter will affect the number of modules identified, and the WGCNA modules provides the  `pickSoftThreshold()` function to help identify good choices for this parameter. 

```{r}
sft <- pickSoftThreshold(normalized_counts,
                         dataIsExpr = TRUE,
                         corFnc = cor,
                         networkType = "signed"
)
sampleTree = hclust(dist(normalized_counts), method = "average")
pdf(file = "Outlier_tree.pdf", width = 12, height = 9);
par(cex = 1.6);
par(mar = c(0,4,2,0))
plot(sampleTree, main = "Sample clustering to detect outliers", sub="", xlab="",
     cex.lab = 1.5,cex.axis = 1.5, cex.main = 2)
dev.off()
```

This `sft` object has a lot of information, we will want to plot some of it to figure out what our `power` soft-threshold should be. 
We have to first calculate a measure of the model fit, the signed $R^2$, and make that a new variable. 

```{r}
# Choose a set of soft threshold parameters
powers = c(c(1:20), seq(from = 22, to=30, by=2))
sft = pickSoftThreshold(normalized_counts, powerVector = powers, verbose = 5) 
# Scale-free topology fit index as a function of the soft-thresholding power
pdf(file = "soft_threshold.pdf", width = 9, height = 5);
par(mfrow = c(1,2));
cex1 = 0.9;
plot(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",type="n",
     main = paste("Scale independence"));
text(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2],
     labels=powers,cex=cex1,col="red");
# this line corresponds to using an R^2 cut-off of h
abline(h=0.90,col="red") 
# Mean connectivity as a function of the soft-thresholding power
plot(sft$fitIndices[,1], sft$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity")) 
text(sft$fitIndices[,1], sft$fitIndices[,5], labels=powers, cex=cex1,col="red")
dev.off()
```

## Turn data expression into topological overlap matrix
```{r}
power=sft$powerEstimate
TOM = TOMsimilarityFromExpr(normalized_counts, power = 12)
dissTOM = 1-TOM 
# Plot gene tree
geneTree = hclust(as.dist(dissTOM), method = "average");
pdf(file = "3-gene_cluster.pdf", width = 12, height = 9);
plot(geneTree, xlab="", sub="", main = "Gene clustering on TOM-based dissimilarity",
     labels = FALSE, hang = 0.04);
dev.off()
```

## Module identification with dynamic cut tree

```{r}
dynamicMods = cutreeDynamic(dendro = geneTree, distM = dissTOM,deepSplit = 2, 
                            pamRespectsDendro = FALSE,minClusterSize = 30);
table(dynamicMods)
length(table(dynamicMods)) 
# Convert numeric labels into colors
dynamicColors = labels2colors(dynamicMods)
table(dynamicColors)
# Plot the dendrogram and colors underneath
pdf(file = "4-module_tree_DGE.pdf", width = 8, height = 6);
plotDendroAndColors(geneTree, dynamicColors, "Dynamic Tree Cut",dendroLabels = FALSE,
                    hang = 0.03,addGuide = TRUE, guideHang = 0.05,main = "Gene dendrogram and module colors")
dev.off()
```

#obtain col data for dds and write the file

```{r}
rowData(dds.raw)
write.csv(as.data.frame(rowData(dds.raw)),file = "dds_norm.csv") #combine the module color data with col data to annotate the modules

#write module colors
write.csv(as.data.frame(dynamicColors),file = "dynamicColors.csv")

#combine the two files to create a module key file

```

#Annotate modules
```{r}

Annotations <- read.csv("Rice_Annotations_Meta.csv")
List <- read.csv("List_dds_norm.csv") #list is created from the dds col1
Annotations_Selected = Annotations [which(Annotations$gene %in% List$gene), ]
write.csv(as.data.frame(Annotations_Selected),file = "MES_annotations_DGE_rice.csv")

#complete the annotation by adding the module colors to the annotations. Use this to determine enrichment using ShinyGO

```


## Merge modules. 

```{r}
MEDissThres=0.5
abline(h=MEDissThres, col = "red")
merge = mergeCloseModules(normalized_counts, dynamicColors, cutHeight = MEDissThres, verbose = 3) 
mergedColors = merge$colors  
mergedMEs = merge$newMEs  
# Plot merged module tree
pdf(file = "5-merged_Module_Tree.pdf", width = 12, height = 9)  
plotDendroAndColors(geneTree, cbind(dynamicColors, mergedColors), 
                    c("Dynamic Tree Cut", "Merged dynamic"), dendroLabels = FALSE, 
                    hang = 0.03, addGuide = TRUE, guideHang = 0.05)  
dev.off()
write.table(merge$oldMEs,file="oldMEs.txt");
write.table(merge$newMEs,file="newMEs.txt")

write.csv(as.data.frame(mergedMEs),file = "mergedMEs_b.csv") #write the module table (merged)

write.csv(as.data.frame(merge$oldMEs),file = "Module_Eingens.csv") #write the module table (unmerged)

write.csv(as.data.frame(merge$newMEs),file = "Module_Eingens_merged.csv") #write the module table (merged)

```

#Get hubgenes
```{R}

colorh = labels2colors(rownames(normalized_counts))
hubs    = chooseTopHubInEachModule(normalized_counts, colorh)
head(hubs)

write.csv(as.data.frame(hubs),file = "HUB_TOP_genes_DGE.csv")
```

#annotate hubgenes

```{r}
Hubs <- read.csv("Rice_Annotations_Meta.csv")
List <- read.csv("List_hub_genes.csv")
Hubs_Selected = Hubs [which(Hubs$gene %in% List$gene), ]
write.csv(as.data.frame(Hubs_Selected),file = "Annotated_Hubs.csv")

```

## Export modules to cytoscape

```{r}
for (i in 1:length(merge$oldMEs)){
  modules = c(substring(names(merge$oldMEs)[i], 3));
  genes = names(as.data.frame(normalized_counts)) ### This is the fix  as.data.frame()
  inModule = is.finite(match(dynamicColors,modules))
  modGenes = genes[inModule]
  modTOM=TOM[inModule,inModule]
  dimnames(modTOM)=list(modGenes,modGenes)
  cyt = exportNetworkToCytoscape(modTOM,
                                 edgeFile = paste("orign_CytoscapeInput-edges-", paste(modules, collapse="-"), ".txt", sep=""),
                                 nodeFile = paste("orign_CytoscapeInput-nodes-", paste(modules, collapse="-"), ".txt", sep=""),
                                 weighted = TRUE, threshold = -1, nodeNames = modGenes, nodeAttr = dynamicColors[inModule]);
}
# Export the gene list of new modules: I still have a problem with this one
for (i in 1:length(merge$newMEs)){
  modules = c(substring(names(merge$newMEs)[i], 3));
  genes = names(as.data.frame(normalized_counts)) ### This is the fix  as.data.frame()
  inModule = is.finite(match(dynamicColors,modules))
  modGenes = genes[inModule]
  modTOM=TOM[inModule,inModule]
  dimnames(modTOM)=list(modGenes,modGenes)
  cyt = exportNetworkToCytoscape(modTOM,
                                 edgeFile = paste("merge_CytoscapeInput-edges-", paste(modules, collapse="-"), ".txt", sep=""),
                                 nodeFile = paste("merge_CytoscapeInput-nodes-", paste(modules, collapse="-"), ".txt", sep=""),
                                 weighted = TRUE, threshold = -1, nodeNames = modGenes, nodeAttr = dynamicColors[inModule]);
}

```



```{r}
# Create the design matrix from the `variety` variable
x.meta <- data.frame(dds$sample, dds$variety) %>%
  set_names(c("sample", "variety")) %>%
  tibble::column_to_rownames(., var = "sample") %>% 
  as.data.frame()

x.meta$variety <- as.factor(x.meta$variety)
des_mat <- model.matrix(data = x.meta, ~variety)# design = ~0+variety

```

Run linear model on each module.
Limma wants our tests to be per row, so we also need to transpose so the eigengenes are rows

```{r}
# lmFit() needs a transposed version of the matrix
fit <- limma::lmFit(t(merge$oldMEs), design = des_mat)

# Apply empirical Bayes to smooth standard errors
fit <- limma::eBayes(fit)
```

Apply multiple testing correction and obtain stats in a data frame. 

```{r}
# Apply multiple testing correction and obtain stats
stats_df <- limma::topTable(fit, number = ncol(merge$oldMEs)) %>%
  tibble::rownames_to_column("module")

#write the stats file
write.csv(as.data.frame(stats_df),file = "stats_df_power_DGE.csv")

```

## Draw a heatmap of pathway enriched for host-pathogen interaction
```{r}

#Heatmap
library(dendsort)
library(gplots)
library(pheatmap)
##read heatmap file;
data_heatmap <- read.csv("Module_Eingens_heatmap_12.csv", comment.char="#")
#write.csv(as.data.frame(module_eigengenes),file = "module_eingens_2.csv")
rnames <- data_heatmap[,1]   # assign labels in column 1 to "rnames"
mat_data <- data.matrix(data_heatmap[,2:ncol(data_heatmap)])  # transform column 2-5 into a matrix
rownames(mat_data) <- rnames

#Cluster rows and columns not needed in RSI
mat_cluster_cols <- hclust(dist(t(mat_data)))
sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...)))
mat_cluster_rows <- sort_hclust(hclust(dist(mat_data)))
mat_cluster_cols <- sort_hclust(mat_cluster_cols)

#Read col and row annotation files
y <- read.table("~/Dropbox/Mac (2)/Documents/WGCNA_Sorghum_19_05/Annotations_Col.txt", header=TRUE)

z <- read.table("~/Dropbox/Mac (2)/Documents/WGCNA_Power_12_Sorghum/Annotation_Rows.txt", header=TRUE)

##plot heatmap
#pdf(file="heatmap_3.pdf")
pheatmap(mat_data, cellwidth = 12, show_rownames= T,show_colnames = T, cluster_rows = F, cluster_cols = mat_cluster_cols, annotation_col = y2,  annotation_row = z, fontsize_row = 8, angle_col = c (90), fontsize_col = 8, col = bluered(120))
#dev.off()





cmd1=cmdscale(as.dist(dissTOM),2)
sizeGrWindow(7, 6)
par(mfrow=c(1,1))
plot(cmd1, col=as.character(color1), main="MDS plot",
xlab="Scaling Dimension 1", ylab="Scaling Dimension 2")


#restGenes= (color1 != "grey")
#diss1=1-TOMsimilarityFromExpr( datExpr[, restGenes], power = 6 )
hier1=hclust(as.dist(diss1), method="average" )
diag(diss1) = NA;
sizeGrWindow(7,7)
TOMplot(diss1^4, hier1, as.character(color),
main = "TOM heatmap plot, module genes" )

power=6
color1=dynamicColors
diss1=1-TOMsimilarityFromExpr(normalized_counts, power = 6 )
hier1=hclust(as.dist(diss1), method="average" )
diag(diss1) = NA;
sizeGrWindow(7,7)
TOMplot(diss1^4, hier1, as.character(color1),
main = "TOM heatmap plot, module genes" )







```

##End

#Runo_May_2022
